---
title: "Behind COVID-19 vaccination target: a SEIR simulation of avoidable deaths and hospitalisations in `r country`"
author: 
  - Lucas Sempe^[Institute of Global Health and Development, Queen Margareth]
  - Aravinda Guntupalli^[Institute of Applied Health Sciences, School of Medicine, Medical Sciences and Nutrition, University of Aberdeen]
  - Peter Lloyd-Sherlock^[School of International Development, University of East Anglia]
output:
  bookdown::word_document2
  toc:false
bibliography: references.bib
---

```{r setup, include=FALSE,echo=F,tinytex.verbose = TRUE, warning=F}

knitr::opts_chunk$set(echo = FALSE, warning = F,
                      message = F, tinytex.verbose = TRUE,
  cache = F,fig.width = 8, dpi=300,
  digits = 1, OutDec = ",")

knitr:: knit_hooks$set(inline = function(x) {
if(is.character(x)){
return(x)
}else{
return(prettyNum(x, big.mark=","))
}
})

```

# Introduction

A year after the launch of the COVID-19 vaccine immunisation process across countries, we note two facts. Vaccines have been proved very effective against existing variants of COVID-19 both in preventing the acquisition of severe disease, hospitalisation and death; and in terms of slowing down the spread of infections[@Imai2021]. Although the way out of the pandemics requires a worldwide solution, we note that the vaccine roll-out has been very unequal across countries. While many high-income nations will have reached high vaccination coverages by the end of 2021 and before the emergence of the highly transmissible Omicron variant [@Thakur2021a], many low- and middle-income countries (LMICs) are still lagging in their vaccination process.

Until the vaccination coverage reaches the vast majority of the worldwide population, there will still be high of uncertainty regarding the future development of the pandemic. Many factors will play a role in local or global outbreaks. Many countries have been already through several epidemic waves, influenced by factors such as the appearance of new variants or the easing of non-pharmaceutical interventions.

On 26 November 2021, the World Health Organization (WHO) designated the coronavirus SARS-CoV-2 B.1.1.529, a variant of concern, named Omicron [@whoUpdateOmicron2021]. As of 22 December 2021, the Omicron variant was already identified in 110 countries across all six WHO Regions [@who1a], becoming predominant across many countries in January 2022 [@hodcroft2022]. On February 9, WHO reported that half a million COVID-19 deaths had been recorded since the Omicron variant was discovered [@AFP2022].

Prior research suggests higher levels of transmission, lower rates of hospitalisation, greater immune evasion and lower vaccine efficacy [@fergusonReport49Growth2021a; @meng2021], although information is still limited to certain countries. This uncertainty significantly increases in LMICs, where civil registration, vital statistics, and epidemiological data is still not robust [@Lloyd-Sherlock2021]. 

While high-income countries are focusing on boosters to protect their citizens, particularly older people, against severe disease, several low and middle-income countries are struggling to provide the first two doses to their vulnerable population. Considering the inequalities within and between countries for the vaccination coverage, In this research, we simulated various scenarios to capture the potential number of deaths and hospitalisations averted in the first semester of 2022 if vaccination levels would be raised to fulfilling WHO’s goal of vaccinating at least 70% of all populations by 1 July 2022. 

# Results

We find that, based on projecting each last month total doses remains constant until July 1st 2022, `r be` countries are behind WHO's goal (see Figure \@ref(fig:mapw)) distributed across the world: `r africa` counties in Africa, `r americas` in the Americas, `r asia` in Asia, `r europe` in Europe and `r oceania` in Oceania. `r islands` countries are small islands with a population less than 600,000 people each. Additionally, `r boost_beh` of these countries already started to provide boosters to their population although they do are behind track in terms of providing two doses for 70% of their population. The number of countries is inferior to the presented by [OWID](https://ourworldindata.org/covid-vaccination-global-projections), as they consider the average number of people who received their first dose of a vaccine per day, over the last 14 days, while our model considers the the average number of people who received any dose of a vaccine per day, over the last 30 days. 

```{r mapw, fig.cap="Countries status according to WHO's vaccination goal based on last month vaccination uptake"}

map_res

```

`r  country` COVID-19 vaccine population coverage in February 8th 2022 is estimated in `r current_coverage100`. This is computed based on the assumption that every person requires two vaccine doses. The comparison between the historical daily vaccination uptake and the needed to reach WHO's target goal for `r  country` is presented in Figure \@ref(fig:vaccines-current). The horizontal lines represent the last month average and the daily needed number of doses. Based on the unvaccinated population and assuming the need of at least two doses per person during the first semester of 2022, we estimate that `r numb` doses are needed to be administered in order to achieve the target of vaccinating 70% of these countries' population. Considering a programmatic delivery cost of US$ 10 per dose [@StrategyAchieveGlobal2021], the estimation reaches `r cost1` american dollars.

```{r vaccines-current, fig.cap="Current and needed daily vaccinations in `r country`"}

plot_vaccines

```

The combination of the different model parameters provide `r scenarios` different scenarios. Figure \@ref(fig:seplot) presents one scenario as an example of the evolution of susceptible over the first semester of 2022: exposed, recovered and deaths, in this case, corresponding to the scenario with the larger number of deaths.

```{r seplot, fig.cap="Number of people in different compartments across time - worst case scenario"}

seplot

```

By comparing a hypothetical scenario without vaccinations and other scenarios where the vaccination coverage reaches 70%, we can estimated that vaccines save between `r dmin0` and `r dmax0` deaths and prevent `r dmin0h` to `r dmax0h` hospitalisations in `r country`, depending on the combinations of parameters.

Simulations where we keep constant the last month vaccination uptake with a parameter R0 = 1.5 across the first semester of 2022 yield an estimated number of deaths ranging from `r min_base` to `r max_base`. When compared to the optimal scenario, we estimate the number of avoidable deaths range from `r dmin` to `r dmax` and the number of avoidable hospitalisations spams from `r dminh` to `r dmaxh`, depending on the combinations of parameters. Table \@ref(tab:tab1) shows the maximum and minimum number of averted deaths and hospitalisations computed across different R0. In the case of the scenarios of averted deaths, the differences between values do not change substantially when R is 1.5 or higher. In the case of the simulation of averted hospitalisations, the minimum values tend to be stable across different, which is explained by the full occupancy of hospital and ICU beds and also the incomplete evolution of the infectious wave due to the period limit imposed to our models.

```{r tab, fig.cap="Maximum and minimum averted deaths and hospitalisations"}

tb %>% kbl (format = "markdown",caption="Maximum and minimum averted deaths and hospitalisations",
            format.args=list(big.mark = ","),digits = 1) 

#
#flextable::flextable(tb)

```

Across the vast majority of models, between 60% and 70% of avoidable deaths correspond to people of 60 years old. This occurs even in majority of countries that currently portray a younger population age structure such cases of regions such as Africa and Asia. Table \@ref(tab:oldtable) summarises the scenarios in terms of total avoidable deaths and infections based on R0 = 1.5 for the population over and under 60 years old. We find that the proportion of avoidable deaths benefits older people, ranging from `r pro1`% to `r pro2`% of the total number of deaths while it is similar across age groups for the total number of hospitalisations, ranging from `r pro3`% to `r pro4`% for the older people group.

```{r oldtable,fig.cap="Proportion of avoidable deaths of older people across simulations"}

old_table %>% kbl (format = "markdown",caption="Proportion of avoidable deaths of older people across simulations",format.args=list(big.mark = ","),digits = 1) 

#flextable::flextable(old_table)

```

Figure \@ref(fig:simulations-ideal) presents three panels with the different simulated scenarios of the number of infections, hospitalisations and deaths averted if WHO's goal of vaccinating 70% of the country population is reached by July 1st 2022. The horizontal axis portrays the different effective reproduction numbers. In the worst case scenario the total number of deaths is estimated to reach `r deaths_averted_max` where `r deaths_averted_older` affect people over 60 years old (`r prop_deaths`% of total) while an intermediate scenario suggests `r deaths_averted_med` deaths.

```{r simulations-ideal, fig.cap="Estimation of infections, hospitalisations and deaths averted based on vaccinating 70% of population by July 1st 2022 - simulated scenarios"}

out_format %>% #filter(max_vaccine>mean_vac)%>%
  group_by(R0,max_vaccine,seeding_cases) %>%
    mutate(`Infections averted`=mean(infections_averted))%>%
    mutate(`Hospitalisations averted`=mean(hospitalisations_averted))%>% 
    mutate(`Deaths averted`=mean(deaths_averted))%>%
pivot_longer(`Infections averted`:`Deaths averted`,names_to = "outcome",values_to = "values") %>%
  mutate(outcome = factor(outcome,levels = c("Infections averted","Hospitalisations averted","Deaths averted"))) %>%
  ggplot() + geom_point(size=3,
    aes(as.factor(R0),values,
                           shape=as.factor(seeding_cases)))+
  theme_light() + facet_wrap(~outcome,scales = "free") + 
  scale_y_continuous(labels = scales::comma) +
  theme(legend.position = "bottom", legend.box="vertical", 
        legend.margin=margin()) + 
  labs(
       caption = "Data source: Authors' own")+
  xlab("Rt") +  ylab("") + scale_shape_manual("Initial number of positive cases",
                                              values = c(15,19,22),
                              labels=c("Official number of cases in last two weeks", "x5","x10"))

```

# Conclusions

These scenarios are built to answer to an ethical framework that aims to find the best possible allocation of COVID-19 vaccines. Our ethical guidelines are the following: we aim to maximise societal health benefits; prioritise those worst-off without the vaccines; and promote equality, where individuals under circumstances shall be treated equally [@persad2020; @emanuel2020]. These principles become operational in terms of saving the most lives; prioritise the most vulnerable populations such as older and immunodeficient people; and protecting health workers. Recently, the WHO SAGE group updated their roadmap for optimal allocation of vaccines across the world, where older adults, health workers, immunocompromised persons, adults with comorbidities, pregnant persons, teachers and other essential workers and disadvantaged  subpopulations at higher risk of severe COVID-19 remain the higher priority groups for additional doses and boosters [@WHO2022].

Vaccines save lives. We have estimated that between `r dmin0` and `r dmax0` deaths can be prevented and between  `r dmin0h` to `r dmax0h` hospitalisations can also be avoided, depending on the combinations of parameters.

Finally, our simulated scenarios suggest the number of deaths could reach `r counter_deaths_max` deaths if the vaccination remains similar to the last month average. Instead, if the vaccination increases to `r n_vac1` dose per day, we estimate `r deaths_averted_max` deaths averted where `r deaths_averted_older` will be people over 60 years old (`r prop_deaths`% of total). The cost associated with the strategy represents `r cost` american dollars.

# References

::: {#refs}
:::

# Empirical strategy

We simulated scenarios based on a previously developed an extended age-structured (5-year groups) stochastic compartmental models of SARS-CoV-2 transmission that includes vaccinations [@Hogan2021; @walker2020]. Older adults age 80 and above were considered in one group due to the smaller sample size in countries with lower life expectancy. The models consider the progression of the population across transmission compartments (susceptible, exposed, infected, recovered), clinical pathways (need for hospitalisation, oxygen and/or intensive care) and vaccination uptake considering factors such as vaccine availability, prioritisation and coverage. The infection transmission model also considers the probability of interactions between age-groups (contact matrices) and the efficacy of the vaccine in terms of prevention of infection and severe disease. 

We employ Susceptible-Exposed-Infectious-Removed (SEIR) models as they rely on an extensive set of parameters such as probabilities of hospitalisation, probability of severe disease, hospital capacity and ICU, reproduction rates, among others. Our simulated scenarios are built based on parameters chosen to represent critical factors affecting the evolution of the pandemics and the vaccination process.

Our outcomes of interest are avoidable hospitalisations and deaths due to vaccination. For that purpose, our research design compares estimations between two scenarios for each country: 

-	a baseline scenario that assumes the currently average daily vaccination level is maintained, and

-	an accelerated scenario where the daily vaccine level is raised to achieve the goal set by the World Health Organisation of vaccinating 70% of the world population against COVID-19 by 1 July 2022 [@StrategyAchieveGlobal2021]. 

For the accelerated scenario, we differentiate between countries that did yet not start providing boosters to their population, and those who already started, even though they are behind meeting the WHO goal.  In the first case, we simulate models with a daily number of vaccines needed to reach two doses for 70% of the population by July 1st 2022, while in the second case, we increase the daily number of vaccines to cover three doses in the same period.

We model a vaccination strategy that prioritises the oldest age group sequentially until the maximum coverage per group is reached. For example, suppose we set the maximum coverage at 90%. In that case, the first age group to be solely vaccinated will be those over 80 and above until it reaches 90% of the age group population. Then, the following group, those aged 75 to 79, will follow the vaccination process followed by 70 to 74, until the youngest group of people get vaccinated. This occurs until the whole eligible population is covered up to 90%. 

As more than 95% of the global procurement of vaccines require two-dose vaccines, our models assume a fully vaccinated person with two vaccines [@imf-who2022; @who2022]. Considering that the models were used only to simulate a single vaccine product, we consider the complexity of multiple vaccine products by weighting the vaccine effectiveness over time [@mrc-ide2022]. Our models assume a dual effect of vaccines' effectiveness in blocking infection and severe disease. We compute the vaccination effectiveness decay for each country and use the world average. This is based on the time since the first vaccination and the time between the first and second doses (we assume 90 days), the decay rate, and different efficacy parameters. Modelling countries that did not start providing boosters, we assume 30% and 60% of infection blocking effectiveness after one and two doses, respectively. In the case of blocking disease, we assume 40% and 80%, respectively. For countries that started applying a booster dose, values assigned are 60% and 80% for infection blocking and 80% and 95% for severe disease blocking. We assume that individuals who have been vaccinated have a 40% reduction in infectiousness if infected. These chosen efficacy values broadly reflect the range of estimated efficacies seen in response to the Omicron variant [@fergusonReport49Growth2021; @fergusonReport50Effectiveness2021; @report1; @khoury2021; @collie2021]. These values do not reflect a specific vaccine as there are unknowns over each specific country vaccine rollout.

The probability of hospitalisation is reduced to 60% in comparison to the prior variants [@fergusonReport50Effectiveness2021]. We assume that the mean duration of naturally acquired immunity is 200 days. 

We also simulate two sets of the number of infected cases: the officially reported number of cases of the last two weeks until 15 February 2022, and this value multiplied by 5 and 10, assuming the lack of massive testing and under-reporting of cases [@lau2021].

The number of people in the first state of the transmission model (those susceptible to the disease) corresponds to the country population. This value is based on the high levels of reinfection found across countries such as South Africa and England [@fergusonReport49Growth2021; @pulliam2021]. We assume a uniform distribution of the vaccinations across adults due to the lack of available data. Recent empirical evidence shows that the generation time for the Omicron variant is shorter than the previous predominant Delta variant [@abbott2022; @liu2021]. Following that, we assume the following parameters: mean duration of 2 days for the incubation period, mean of 2.6 days for a mild infection and a mean of 3.8 days for symptoms onset to admission to hospital.

The time period for the analysis is 1 year, from 15 February 2022 to 14 February 2023. This allows to capture both the date set by the World Health Organisation’s target to vaccinate 70% of the world’s population against COVID-19 by mid-2022 [@StrategyAchieveGlobal2021] and the decay rate of vaccine effectiveness across time.

We use a constant effective reproduction number of 1.3, 1.5 and 2 [@vanwees2021; @huang2022; @ignatov2022] across the first half of 2022. In the second part of the year, the reproduction number is set in .9, which implies the contraction of the pandemics over time.

Table \@ref(tab:param1) provides a summary of all the main parameters. Additional country-specific, epidemiological and vaccination parameters were compiled by @Hogan2021 and updated in the R package ['nimue'](https://github.com/mrc-ide/nimue), where original sources of data are given. Infections and vaccine data are collected from Our World in Data [@roser2020]. Probabilities of death according different states and age groups (severity of disease, treatment and information source) can be found in the Appendix.


```{r param1}

Parameters <- read_excel("Parameters.xlsx")

Parameters %>% kbl (format = "markdown",caption="Parameters used across simulation scenarios based on Omicron variant",
                    format.args=list(big.mark = ","),digits = 1) 


```
