---
title: "prueba"
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The next phase of the pandemics: who will bear the brunt?

Goal is to estimate/simulate projected infections and fatalities amongst older persons in LMIC and the (lack of) impact of vaccination uptake.

## Assumptions and limitations

1)  Data in LMICs is still not robust, especially in terms of cases due to testing capacity and quality.

2)  Although we know important facts such as the role of vaccination towards mitigating infections, there is a lot of uncertainty of future due to diverse factors such as the raise of new variants, the seasonality of the infection, geographical outbreaks, among others.

3)  Simulating scenarios with a reasonable range of parameters appears to be a robust way of projecting the future development of the pandemic.

## Empirical strategy

### Data sources

a) We use the Cover-Age-DB database, which offers a standardized count measures by sex and harmonized age groups across the world [@riffe2020].

b) We also use a global excess mortality database [@karlinsky2021].

c) Vaccine uptake (per group age). Explore national databases. If not, work with scenarios

### Methods

The simulations are build in three steps.

1) ***We select countries where the under estimation of the mortality ratio is between the range of .9 and 1.1***. This means that only countries where the current data reflects relatively well excess mortality are taken account for.

2) We estimate Age standardised case fatality rates (CFR) per country. We compare with neighbouring countries to asses if lack of testing may produce bias. 

For example, in case of comparing case fatality rates from Peru, Colombia and Mexico, we observe that CFRs are significantly higher in Peru, which is even higher for older people. 

***We discard countries with CFRs that differ significantly from their neighbours (alternative = use a regional CFR)***

```{r}

per  %>% group_by (Age) %>% filter (!is.na(Deaths)) %>% 
  mutate (Age = as.factor(Age)) %>%
  mutate (CFR = Deaths / Cases) %>% 
  ggplot ( ) + 
  geom_hline (yintercept = 1, 
                         linetype = 2, 
                         colour="darkorange") +
  geom_line (aes(Date, CFR, colour=Age)) + 
  theme_classic ( ) + facet_wrap (~Country) + 
  ylim (0,2) + theme(legend.position = "bottom")
  
```

3) We use those parameters (CFR and deaths)  + vaccination uptake to simulate SEEIR models with vaccine 






References
